Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2025-01-02 manuscript.Rmd
update results and figure to include new analysis

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation biomass and soil, and this carbon storage capacity depends on the type of the AFS in addition to environmental and edaphic conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable: although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e., planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFS, while AFS such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate and previous land use type can also influence SOC in AFS, which may explain the variability of observations reported in the literature [@Feliciano2018; @Ma2020; @Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS was only evidenced to a depth of 10 cm [@Ma2020].

@@

[paragraph on possible drivers]
C stocks (in biomass and soil) of AFS could be driven by a number of factors: 
- agricultural practices (ref)
- climate (ref)
- soil properties (especially for SOC) > in particular, fine content

@@

```{r get-env-data}
# temperature and precipitation from CHELSA
if (!file.exists("data/CHELSA_bio1_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio1_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio1_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

@@

if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio12_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio12_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

@@

# 30m resolution SRTM
if (!file.exists("data/srtm_62_14.zip")) {
  url <- "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_62_14.zip"
  destfile <- "data/srtm_62_14.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}

# download cwd from Chave et al 2014
if (!file.exists("data/CWD.tif")) {
  url <- "http://chave.ups-tlse.fr/pantropical_allometry/CWD.tif.zip"
  destfile <- "data/CWD.tif.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}
```

```{r extract-env-data}
# extract elevation, precipitation and temperature for all plots
data_plot[, `:=`(
  elev = terra::extract(terra::rast("data/srtm_62_14.tif"), 
                        cbind(long, lat))[[1]], 
  prec = terra::extract(terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  temp = terra::extract(terra::rast("data/CHELSA_bio1_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]]
)]

cor(data_plot[, c("elev", "prec", "temp")])

@@

We used the typology of AFS described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair *et al.* (1993, 2021) and adapted to the local context with semi-structured interviews previously conducted in another study at the same sites [@Cogne2024]. The five AFS identified were as follows: Crop systems including a wooded Fallow phase (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). The AFS are presented in more detail in the supplemental information section (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

## Field inventories

@@

Taxonomic identification was carried out using Tetum (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory Government's Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots (including the plot center and two opposite corners; Figure S\@ref(fig:design)), covering the intra-plot variability and at a minimum distance of 1 m from any tree. Soils were sampled with a soil auger ( Model Bor Tanah - Dormer Standard Soil Auger, Adiguna Karya Persada, Bogor, Indonesia) at 0-10 cm and 20-30 cm depth,. As the length of the head auger (15 cm) was longer than our target soil depths (10 cm), we could not accurately sample the 10-20 cm soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). Total carbon content was determined by dry combustion, following the ISO 10694:1995 standard method, and measured with a thermal conductive detector (EA 1112 Elemental Analyser, Thermo Fischer Scientific, Netherlands). Carbonate content was determined with a calcimeter, following the ISO 10693 standard method. Soil organic carbon content was calculated as the difference of the total carbon and the carbonate contents.

@@

```{r bgc-estimation}
# root-to-shoot ratio allometry from Ledo et al., 2018
# use CWD values from Chave et al., 2014
data_plot[, cwd := terra::extract(terra::rast("data/CWD.tif"), cbind(long, lat))]
data_sim_agc <- merge(data_sim_agc, data_plot[, c("N_PLOT", "cwd")], by = "N_PLOT")

# add uncertainty on R:S ratio? 
data_sim_agc[, RS := exp(- 1.2312 - 0.0215 * DIAM_130_CM + 0.0002 * DIAM_130_CM^2 - 0.0007 * cwd)]
data_sim_agc[, BGC := RS * AGC]
```

@@

Where $Corg_i$ is the SOC content at depth i (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger and estimated at depth 10-20 cm (see details below); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth i (in g), converted into Mg with a factor of 10$^-6$; $Surf\ =\ 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method M4 recommended by @Poeplau2017, which correctly considers the coarse fraction of the soil. We considered that $mfine_i$ at depths of 2-7 cm, 12-17 cm and 22-27 cm was associated with $Corg_i$ at depths of 0-10, 10-20 cm and 20-30 cm, respectively. At depth $i$ = 10-20 cm, where the carbon content was not measured (see “Soil measurements” section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a standard negative exponential model of SOC content decrease with soil depth [@Mishra2009;@Murphy2019]:

@@

where $Corg_{a,b}$ is the average SOC content between depths a and b (see rationale for equation \@ref(eq:intersoc) in Supplementary Information xx).

@@

```{r add-other-soil-properties}
data_soil_summ <- data_soil |> 
  subset(!is.na(TC)) |> 
  melt(id.vars = c("N_PLOT", "SAMPLE_DEPTH"), 
       measure.vars= c("CLAY", "FINE SILT", "pH\n", "Ca", "CEC\n"))  |> 
  setDT()
data_soil_summ[, variable := gsub("\n", "", variable)]
data_soil_summ <- 
  data_soil_summ[, .(value = mean(as.numeric(value), na.rm = TRUE)), .(N_PLOT, variable)] |> 
  dcast(N_PLOT ~ variable)
data_soil_summ[, fine_soil := CLAY + `FINE SILT`] 
data_sim <- merge(data_sim, data_soil_summ)  
# add elevation
data_sim <- merge(data_sim, data_plot[, c("N_PLOT", "elev")])
```

@@

We modelled both AGC and SOC stocks as:

$$Y_{i,j,k} = \mu +  \theta \cdot X_i + \alpha_{j} + \beta_{k} + \epsilon_{i}$$
with $Y_{i,j,k}$ the stocks (either SOC or AGC) of plot $i$ in site $k$ and AFS type $j$

@@

$\theta$ is the effect of the covariates on $Y$, and $X_i$ are the values of the covariates in plot $i$; they include elevation for AGC stocks [from NASA SRTM; @SRTM2024], or elevation [@SRTM2024], clay + fine soil content (from soil analysis; xxx methods) and AGC stocks for SOC stocks. All covariates were centred and scaled. Elevation was chosen as it was highly correlated (>95%) with both mean annual temperature and precipitation [@Karger2017]: it was thus a good indicator of a climate gradient in our dataset (see Supplementary information section xx). xxx justify these variables in the introduction

@@

```{r}
stan_model <- "
data {
  int<lower=0> N;
  int<lower=1> AF;
  int<lower=1> S;
  int<lower=1> K;
  int<lower=1,upper=AF> afs[N];
  int<lower=1,upper=S> site[N];
  vector[N] elev;
  matrix[N,K] covar;
  real<lower=0> agc[N];
  real<lower=0> soc[N];
  real<lower=0> soc10[N];
}

parameters {
  real mu_a;
  real delta_afs_a[AF];
  real delta_site_a[S];
  real theta_a;
  real<lower=0> sigma_a;

  real mu_s;
  real delta_afs_s[AF];
  real delta_site_s[S];
  vector[K] theta_s;
  real<lower=0> sigma_s;

  real mu_s10;
  real delta_afs_s10[AF];
  real delta_site_s10[S];
  vector[K] theta_s10;
  real<lower=0> sigma_s10;

@@

  // predictions
  for (i in 1:N) {
    pred_a[i] = mu_a + elev[i] * theta_a + delta_afs_a[afs[i]] + delta_site_a[site[i]];
    pred_s[i] = mu_s + dot_product(row(covar, i), theta_s) + delta_afs_s[afs[i]] + delta_site_s[site[i]];
    pred_s10[i] = mu_s10 + dot_product(row(covar, i), theta_s10) + delta_afs_s10[afs[i]] + delta_site_s10[site[i]];
  }

  // likelihoods
  agc ~ normal(pred_a, sigma_a);
  soc ~ normal(pred_s, sigma_s);
  soc10 ~ normal(pred_s10, sigma_s);

@@

  //  priors
  theta_s ~ normal(0,1);
  theta_s10 ~ normal(0,1);
  sigma_a ~ normal(0,1);
  sigma_s ~ normal(0,1);
  sigma_s10 ~ normal(0,1);
  sigma_site_a ~ normal(0,1);
  sigma_afs_a ~ normal(0,1);
  sigma_site_s ~ normal(0,1);
  sigma_afs_s ~ normal(0,1);
  sigma_site_s10 ~ normal(0,1);
  sigma_afs_s10 ~ normal(0,1);
}

@@

```{r stan}
if (!dir.exists("outputs")) dir.create("outputs")

@@

      data_stan <- data_sim |> 
        subset(iter==i) |> 
        dcast(N_PLOT + site + ID_AF + fine_soil + elev ~ variable) |> 
        with(
          list(
            N = length(N_PLOT),
            AF = length(unique(ID_AF)),
            S = length(unique(site)),
            K = 3,
            afs = as.numeric(as.factor(ID_AF)), 
            site = as.numeric(as.factor(site)),
            elev = (elev-mean(elev))/sd(elev),
            covar = scale(cbind(elev, fine_soil, AGC)),
            agc = AGC,
            soc = SOC,
            soc10 = SOC10
          ))

@@

data_par[, level_num := as.numeric(gsub("V", "", level_num))]
data_par[effect == "afs" & !is.na(level_num), 
         level := levels(as.factor(data_sim$ID_AF))[level_num]]
data_par[effect == "site" & !is.na(level_num),
         level := levels(as.factor(data_sim$site))[level_num]]
data_par[param == "theta",
         level := c("elev", "fine", "agc")[level_num]]
data_par[param == "theta" & stock == "agc", level := "elev"]

@@

@@

## Effect of elevation and fine soil on AGC and SOC stocks

@@

The effect of AGC on SOC was close to zero (Figure \@ref(fig:plot-reg)).

@@

```{r fig.id = "effect-afs-site", fig.width = 7, fig.height = 3}
labels_stock = function(x) paste(toupper(x), "(Mg/ha)")
labels_effect = function(x) ifelse(x %in% c("afs", "agc"), toupper(x), x)

@@

```{r fig.id = "effect-agc-clay", fig.width = 7, fig.height = 2}
labels_level = function(x)
  ifelse(x=="agc", "AGC", ifelse(x == "elev", "Elevation", "Clay + fine silt"))
pars_summary |> 
  subset(param == "theta")  |> 
  ggplot(aes(y = ifelse(stock=="soc10", "SOC : 0-10 cm", 
                        ifelse(stock == "soc", "SOC : 0-30 cm", "AGC")))) + 
  geom_vline(xintercept = 0, lty = 2) +
  geom_segment(aes(x = low95, xend = upp95)) + 
  geom_segment(aes(x = low80, xend = upp80), col = 2, lwd = 3) + 
  geom_point(aes(x = mean), size = 2) +
  facet_wrap(~level, labeller = labeller(level = labels_level)) + 
  labs(x="Effect on SOC", y = NULL) +
  theme_bw()
```

```{r fig.id = "plot-reg", fig.height = 3.5, fig.width = 7, fig.cap = "Scatter plot of aboveground carbon (AGC; x-axis) and soil organic carbon (SOC; y-axis) stocks. SOC stocks were estimated at the topsoil layer (0-10 cm; left panel) and at the total measured depth (0-30 cm; right panel). Each point represents the median C stock value in a sample plot, and the vertical and horizontal error bars are the 95% confidence interval. The colors correspond to the different agroforestry systems (CF: crop and fallow, SP: silvopasture, YA: young agroforest, HG: home garden, FG: forest garden); the different shapes represent the two sites. The dotted lines represent the (non-significant) linear mixed models between log-transformed AGC and SOC stocks. For each linear mixed model, we report the marginal R^2^ (proportion of variance explained by fixed effects only) and the conditional R^2^ (proportion of variance explained by both fixed and random effects). Both x- and y-axes are log-transformed. One plot (AFS: CF, site: Gariuai) with no trees over 30 cm in circumference at breast height (minimum size for measurement) had AGC stocks estimated at 0 Mg C ha^-1^ and was therefore removed from the linear model and the figure."}

@@

# Supplementary information

## Supplementary methods

@@

$$ \Rightarrow C_{10,20} = C_{0,10} \cdot \left( \frac{C_{20,30}}{C_{0,10}}\right)^{\frac{1}{2}}$$

@@

>>>>

Agroforestry systems (AFS) have the potential to store large amounts of carbon in the vegetation biomass and soil, and this carbon storage capacity depends on the type of the AFS in addition to environmental and edaphic conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable: although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e., planting trees with intercropping during the establishment phase) may decrease SOC stocks compared to non-AFS, while AFS such as silvopastures have been shown to strongly increase SOC stocks [@Feliciano2018]. Other factors such as climate, soil characteristics and previous land use type can also influence SOC in AFS, which may explain the variability of observations reported in the literature [@Feliciano2018; @Ma2020; @Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS was only evidenced to a depth of 10 cm [@Ma2020].

@@

@@

```{r get-env-data}
# mean annual temperature from CHELSA
if (!file.exists("data/CHELSA_bio1_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio1_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio1_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

@@

# mean annual precipitation from CHELSA
if (!file.exists("data/CHELSA_bio12_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio12_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio12_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

@@

# precipitation variability from CHELSA
if (!file.exists("data/CHELSA_bio15_1981-2010_V.2.1.tif")) {
  url <- "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/climatologies/1981-2010/bio/CHELSA_bio15_1981-2010_V.2.1.tif"
  destfile <- "data/CHELSA_bio15_1981-2010_V.2.1.tif"
  download.file(url, destfile, method = "curl")

# download cwd from Chave et al 2014
if (!file.exists("data/CWD.tif")) {
  url <- "http://chave.ups-tlse.fr/pantropical_allometry/CWD.tif.zip"
  destfile <- "data/CWD.tif.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
}

# 30m resolution SRTM
if (!file.exists("data/srtm_62_14.zip")) {
  url <- "https://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_5x5/TIFF/srtm_62_14.zip"
  destfile <- "data/srtm_62_14.zip"
  download.file(url, destfile, method = "curl")
  unzip(destfile, exdir = "data", overwrite = TRUE)
} 
```

```{r extract-env-data}
# extract elevation, precipitation and temperature for all plots
data_plot[, `:=`(
  elev = terra::extract(terra::rast("data/srtm_62_14.tif"), 
                        cbind(long, lat))[[1]], 
  temp = terra::extract(terra::rast("data/CHELSA_bio1_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  prec = terra::extract(terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  varp = terra::extract(terra::rast("data/CHELSA_bio15_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  cwd = terra::extract(terra::rast("data/CWD.tif"), 
                        cbind(long, lat))[[1]]
)]

@@

We used the typology of AFS described in @Cogne2024, which was developed in the same study area. This typology was based on the one proposed by Nair *et al.* (1993, 2021) and adapted to the local context with semi-structured interviews previously conducted in another study at the same sites [@Cogne2024]. The five AFS identified were as follows: Crop systems including a wooded Fallow phase (CF), Silvopastures (SP), Young Agroforests (YA), Home Gardens (HG), and Forest Gardens (FG). The AFS are presented in more detail in the Supplemental Information (Table S\@ref(tab:afs-typology)) and in @Cogne2024.

@@

## Field inventories {#field-inventories}

@@

Taxonomic identification was carried out using Tetum (a national official language) vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory Government's Native Plants Herbarium [@NT2013]. The correspondence between Tetum names and scientific names is provided in the Supplemental Information (Table S\@ref(tab:tetum-names)).

In each of the 30 sample plots, a composite of three samples was taken from three of the five 10 x 10 m subplots (including the plot center and two opposite corners; Figure S\@ref(fig:design)), covering the intra-plot variability and at a minimum distance of 1 m from any tree. Soils were sampled with a soil auger (Model Bor Tanah - Dormer Standard Soil Auger, Adiguna Karya Persada, Bogor, Indonesia) at 0-10 cm and 20-30 cm depth. As the length of the head auger (15 cm) was longer than our target soil depths (10 cm), we could not accurately sample the 10-20 cm soil layer. Soils samples were then air-dried and sieved at 2 mm. They underwent standard physical and chemical analyses at Cirad laboratories in Montpellier, France (Table S\@ref(tab:site-soil)). Total carbon content was determined by dry combustion, following the ISO 10694:1995 standard method, and measured with a thermal conductive detector (EA 1112 Elemental Analyser, Thermo Fischer Scientific, Netherlands). Carbonate content was determined with a calcimeter, following the ISO 10693 standard method. Soil organic carbon content was calculated as the difference of the total carbon and the carbonate contents. Soil texture (proportion of clay, silt and sand), pH, xxx and CEC were measured as xxx.

@@

```{r bgc-estimation}
# root-to-shoot ratio allometry from Ledo et al., 2018
# use CWD values from Chave et al., 2014

data_sim_agc[, RS := exp(- 1.2312 - 0.0215 * DIAM_130_CM + 0.0002 * DIAM_130_CM^2 - 0.0007 * cwd)]
data_sim_agc[, BGC := RS * AGC]
```

@@

Where $Corg_i$ is the SOC content at depth i (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger and estimated at depth 10-20 cm (see details below); $mfine_i$ is the mass of the fine fraction of the soil in the sampling cylinder at depth i (in g), converted into Mg with a factor of 10$^-6$; $Surf\ =\ 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm) and $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$). This calculation corresponds to the method M4 recommended by @Poeplau2017, which correctly considers the coarse fraction of the soil. We considered that $mfine_i$ at depths of 2-7 cm, 12-17 cm and 22-27 cm was associated with $Corg_i$ at depths of 0-10, 10-20 cm and 20-30 cm, respectively. At depth $i$ = 10-20 cm, where the carbon content was not measured (see “Field inventories” section), we estimated SOC content as a function of SOC content at 0-10 cm and 20-30 cm, assuming a standard negative exponential model of SOC content decrease with soil depth [@Mishra2009;@Murphy2019]:

@@

where $Corg_{a,b}$ is the average SOC content between depths a and b (see rationale for equation \@ref(eq:intersoc) in Supplemental Information).

@@

```{r add-other-soil-properties}
data_soil_summ <- data_soil |> 
  subset(!is.na(TC)) |> 
  melt(id.vars = c("N_PLOT", "SAMPLE_DEPTH"), 
       measure.vars= c("CLAY", "FINE SILT", "pH\n", "Ca", "CEC\n"))  |> 
  setDT()
data_soil_summ[, variable := gsub("\n", "", variable)]
data_soil_summ <- 
  data_soil_summ[, .(value = mean(as.numeric(value), na.rm = TRUE)), .(N_PLOT, variable)] |> 
  dcast(N_PLOT ~ variable) 
data_sim <- merge(data_sim, data_soil_summ)  
# add mean annual temperature
data_sim <- merge(data_sim, data_plot[, c("N_PLOT", "temp")])
```

@@

### Carbon stocks model

To test the relative effect of AFS type, climate and soil on carbon stocks, we modelled both AGC and SOC stocks as:

$$Y_{i,j,k} = \mu +  \sum_{l} \left( \theta_l \cdot X_{l,i} \right) + \alpha_{j} + \beta_{k} + \epsilon_{i}$$
with $Y_{i,j,k}$ the stocks (either AGC or SOC) of plot $i$ in site $k$ and AFS type $j$.

@@

$\theta_l$ is the effect of covariate $l$ on variable $Y$, and $X_{l,i}$ is the value of covariate $l$ in plot $i$. We included the following covariates: elevation [from NASA SRTM; @SRTM2024] to represent the climate gradient in our study area, and clay fraction (as measured in the Cirad laboratories; see “Field inventories” section) to represent the soil gradient in our study area. The rationale for the choice of these variables is given in the Supplemental Information. In addition, we included AGC stock as a covariate in the SOC stock model, as we wanted to quantify its effect on SOC stocks (hypotheses H2a and H2b). Covariates were all centred and scaled.

@@

```{r stan-model}
stan_model <- "
data {
  int<lower=0> N;
  int<lower=1> AF;
  int<lower=1> S;
  int<lower=1> KA;
  int<lower=1> KS;
  int<lower=1,upper=AF> afs[N];
  int<lower=1,upper=S> site[N];
  matrix[N,KA] covarA;
  matrix[N,KS] covarS;
  real<lower=0> agc[N];
  real<lower=0> soc[N];
  real<lower=0> soc10[N];
}

parameters {
  real mu_a;
  real delta_afs_a[AF];
  real delta_site_a[S];
  vector[KA] theta_a;
  real<lower=0> sigma_a;

  real mu_s;
  real delta_afs_s[AF];
  real delta_site_s[S];
  vector[KS] theta_s;
  real<lower=0> sigma_s;

  real mu_s10;
  real delta_afs_s10[AF];
  real delta_site_s10[S];
  vector[KS] theta_s10;
  real<lower=0> sigma_s10;

@@

  // predictions
  for (i in 1:N) {
    pred_a[i] = mu_a + dot_product(row(covarA, i), theta_a) + delta_afs_a[afs[i]] + delta_site_a[site[i]];
    pred_s[i] = mu_s + dot_product(row(covarS, i), theta_s) + delta_afs_s[afs[i]] + delta_site_s[site[i]];
    pred_s10[i] = mu_s10 + dot_product(row(covarS, i), theta_s10) + delta_afs_s10[afs[i]] + delta_site_s10[site[i]];
  }

  // likelihoods
  agc ~ normal(pred_a, sigma_a);
  soc ~ normal(pred_s, sigma_s);
  soc10 ~ normal(pred_s10, sigma_s10);

@@

  //  priors
  theta_a ~ normal(0,1);
  sigma_a ~ normal(0,1);
  sigma_site_a ~ normal(0,1);
  sigma_afs_a ~ normal(0,1);
  theta_s ~ normal(0,1);
  sigma_s ~ normal(0,1);
  sigma_site_s ~ normal(0,1);
  sigma_afs_s ~ normal(0,1);
  theta_s10 ~ normal(0,1);
  sigma_s10 ~ normal(0,1);
  sigma_site_s10 ~ normal(0,1);
  sigma_afs_s10 ~ normal(0,1);
}

@@

```{r stan-calibration}
if (!dir.exists("outputs")) dir.create("outputs")

@@

      data_stan <- data_sim |> 
        subset(iter==i) |> 
        dcast(N_PLOT + site + ID_AF + CLAY + temp ~ variable) |> 
        with(
          list(
            N = length(N_PLOT),
            AF = length(unique(ID_AF)),
            S = length(unique(site)),
            KA = 2,
            KS = 3,
            afs = as.numeric(as.factor(ID_AF)), 
            site = as.numeric(as.factor(site)),
            covarA = scale(cbind(temp, CLAY)),
            covarS = scale(cbind(temp, CLAY, AGC)),
            agc = AGC,
            soc = SOC,
            soc10 = SOC10
          ))

@@

data_par[, level_num := as.numeric(gsub("V", "", level_num))]
data_par[effect == "afs" & !is.na(level_num), 
         level := levels(as.factor(data_sim$ID_AF))[level_num]]
data_par[effect == "site" & !is.na(level_num),
         level := levels(as.factor(data_sim$site))[level_num]]
data_par[param == "theta",
         level := c("mat", "clay", "agc")[level_num]]

@@

## Distribution of carbon stocks by AFS and site

@@

## Effect of mean annual temperature and clay fraction on AGC and SOC stocks

Zero was included in the 80% credibility intervals for the effect of mean annual temperature and clay fraction on AGC, meaning that no strong effect was detected in our data (Figure \@ref(fig:effect-agc-clay)). Similarly, the clay fraction did not show a strong effect on SOC stocks, as zero was included in its 80% credible intervals (Figure \@ref(fig:effect-agc-clay)).

Mean annual temperature, on the contrary, had a stronger negative effect on SOC stocks at both 0-10 cm and 0-30 cm depth (Figure \@ref(fig:effect-agc-clay)). Values of $\theta_{MAT}$ were negative in `r sum(subset(data_par, stock=="soc" & param == "theta" & level == "mat")$value<0)/10`% of cases for SOC at 0-30 cm depth, and in `r sum(subset(data_par, stock=="soc10" & param == "theta" & level == "mat")$value<0)/10`% of cases for SOC at 0-10 cm depth. They averaged `r round(mean(subset(data_par, stock=="soc" & param == "theta" & level == "mat")$value), 2)` at 0-30 cm depth, meaning that SOC stocks were estimated to decrease by `r abs(round(mean(subset(data_par, stock=="soc" & param == "theta" & level == "mat")$value), 2))` Mg C ha$^{-1}$ for an increase in mean annual temperature of one standard deviation, i.e. `r round(sd(data_plot$temp), 2)` °C.

@@

The relationship between AGC and SOC stocks was weak in our dataset (Figure  \@ref(fig:plot-reg)), as the 80% credibility interval of the associated parameter ($\theta_{AGC}$) included zero (Figure \@ref(fig:effect-agc-clay)).

@@

```{r fig.id = "effect-afs-site", fig.width = 7, fig.height = 3, fig.cap = "Effect of AFS (CF: crop and fallow, SP: silvopasture, YA: young agroforest, HG: home garden, FG: forest garden) and site on carbon stocks (in MgC/ha). These values were obtained as the sum of the mean predicted stocks ($\\mu$) and the AFS or site random effects. The black dot represents the mean of the 1000 iterations; the red segment represents the 80% confidence interval and the black segment represents the 95% confidence interval. AGC: aboveground (biomass) carbon; SOC: soil organic carbon."}
labels_stock = function(x) paste(toupper(x), "(Mg/ha)")
labels_effect = function(x) ifelse(x %in% c("afs", "agc"), toupper(x), x)

@@

```{r fig.id = "effect-agc-clay", fig.width = 7, fig.height = 2, fig.cap = "Effect of covariates (MAT: mean annual temperature, Clay: clay fraction, AGC: aboveground carbon) on carbon stocks (SOC: soil organic carbon, at 0-10 cm or 0-30 cm depth). Covariates are all centred and scaled; positive values indicate that C stocks tend to be higher for higher values of the covariate. Black dots represent the mean of the 1000 iterations; red segments represent 80% confidence intervals and black segments represent 95% confidence intervals."}
labels_stock = function(x)
  ifelse(x=="agc", "AGC", ifelse(grepl("10", x), "SOC (0-10 cm)", "SOC (0-30 cm)"))

pars_summary |> 
  subset(param == "theta")  |> 
  ggplot(aes(y = level)) + 
  geom_vline(xintercept = 0, lty = 2) +
  geom_segment(aes(x = low95, xend = upp95)) + 
  geom_segment(aes(x = low80, xend = upp80), col = 2, lwd = 3) + 
  geom_point(aes(x = mean), size = 2) +
  facet_wrap(~stock, labeller = labeller(stock = labels_stock)) + 
  labs(x="Effect of covariates on C stocks", y = NULL) +
  scale_y_discrete(labels= c("AGC", "Clay", "MAT")) +
  theme_bw()
```

```{r fig.id = "plot-reg", fig.height = 3.5, fig.width = 7, fig.cap = "Scatter plot of aboveground carbon (AGC; x-axis) and soil organic carbon (SOC; y-axis) stocks. SOC stocks were estimated at the topsoil layer (0-10 cm; left panel) and at the total measured depth (0-30 cm; right panel). Each point represents the mean SOC stocks in a sample plot, and the vertical and horizontal error bars are the 95% confidence intervals. The colours correspond to the different agroforestry systems (CF: crop and fallow, SP: silvopasture, YA: young agroforest, HG: home garden, FG: forest garden); the different shapes represent the two sites. The dotted lines represent the mean predicted SOC stocks for a given AGC stock value, and the grey areas represent the 95% confidence interval of this prediction."}

@@

# Supplemental Information

## Supplemental methods

@@

$$ \Rightarrow C_{10,20} = C_{0,10} \cdot \left( \frac{C_{20,30}}{C_{0,10}}\right)^{\frac{1}{2}}$$
### Choice of climate and soil covariates

As described in the "Carbon stocks model" section of the Materials and Methods, we hypothesized that C stocks (both AGC and SOC) are affected by climate and soil characteristics, together with land use (in this case, AFS type) and other spatial differences (in our case, a site effect).

```{r pca-clim-soil}
pca_clim <- data_plot[, c("temp", "prec", "varp")] |> 
  scale() |> princomp()

data_soil_melt <- 
  data_soil |> 
  melt(id.vars = c("N_PLOT", "SAMPLE_DEPTH"), 
       measure.vars = c(
         "pH\n", "CEC\n",  "CLAY", "FINE SILT", 
         "COARSE SILT", "FINE SAND", "COARSE SAND"))

# make values numeric; when detectability limit is reached, used limit value
data_soil_melt[grepl("LD", value), value := gsub(",", ".", tstrsplit(value, "LD=")[[2]])]

data_soil_melt <- 
  data_soil_melt[, .(value = mean(as.numeric(value), na.rm = TRUE)), .(N_PLOT, variable)] |> 
  dcast(N_PLOT ~ variable) 
data_soil_melt[, silt := `COARSE SILT` +`FINE SILT`]
data_soil_melt[, sand := `FINE SAND` + `COARSE SAND`]

pca_soil <- data_soil_melt[, c("pH\n", "CEC\n",  "CLAY", "silt", "sand")] |> 
  scale() |> princomp()
```

We selected the following set of variables to represent the effect of climate: mean annual temperature (BIO1), mean annual precipitation (BIO12), annual precipitation variability (BIO15) over the period 1981-2010. These bioclimatic variables were extracted from CHELSA 2.1 [@Krager2017]. These variables were all highly correlated in our dataset (Figure S\@ref(fig:choice-clim-vars)a). We performed a principal component analysis after scaling and centering these three variables: the first axis explained `r round(pca_clim$sdev[1]^2/sum(pca_clim$sdev^2)*100)`% of the variance in our dataset (Figure S\@ref(fig:choice-clim-vars)b). We therefore decided to keep only one variable to represent this climatic gradient: mean annual temperature, which was most correlated with the first axis of the PCA (Figure S\@ref(fig:choice-clim-vars)b), and has been shown to influence C stocks in other contexts [@Wiesmeier2019;@Muller-Landau2021].

We selected the following set of variables to represent the effect of soil properties on C stocks: clay fraction (<2 $\mu$m), silt fraction (2-50 $\mu$m), sand fraction (50-2000 $\mu$m), pH and cation exchange capacity xx. These edaphic variables were measured on our soil samples analysed in the Cirad laboratories (see section "Field inventories"). After scaling and centering these variables, we performed a principal component analysis: the first axis explained `r round(pca_soil$sdev[1]^2/sum(pca_soil$sdev^2)*100)`% of the variance in our data set (Figure S\@ref(fig:choice-soil-vars)b). We therefore decided to retain only one variable to represent this edaphic gradient: clay fraction, which was one of the most highly correlated with the first axis of the PCA (Figure S\@ref(fig:choice-soil-vars)b) and has been shown to influence C stocks in other contexts [@Wiesmeier2019].

@@

<!-- </br> -->
